Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-04-08 manuscript/manuscript.qmd
updated introduction

#### Limitations of Existing Approaches
Regrettably, the development process for traditional ocean color algorithms is founded on a statistical erroneous argument. The fallacy is the error of conflating empirical frequency — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials,Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena. These samples' probabilities are considered the observed frequencies that capture the true - albeit noised up and potentially biased - behavior of the system. Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $M$ is true. The fallacy arises from the mistaken belief that, with sufficiently large amounts of data, one can simply "flip" the likelihood to obtain $p(M∣D)$, thereby yielding a probabilistic statement about $M$. Compounding this error is the imposition of numerous unverified assumptions—disguised as objectivity—that lead frequentists to set the parameters of $M$ by maximizing $p(D∣M)$, under the false premise that this provides the best possible representation of the model. In essence, this fallacy leads to a confusion between the sampling probability $p(D∣M)$ and the inferential probability $p(M∣D)$; the latter  when correctly obtained  quantifies the plausibility of the hypothesis represented by $M$ - precisely what we seek. The historical dimensions of this frequentist illusion are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

Regrettably, the development of traditional ocean color algorithms relies on a fundamental statistical error. The error is the mistaken assumption that the relative frequencies observed in a finite sample can be taken as precise estimates of the underlying true probabilities. In other words, it is the error of conflating the empirical frequency—derived from a limited, noisy set of observations—with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling. The fallout has and continues to be extremely detrimental. Traditional, frequentist approaches, where the variability in sample frequencies is erroneously treated as negligible, yield overconfident inferences, and allegedgly significant findings that do not hold water, in time leading to replication crises.

In the context of statistical modelling, where $M$ is a model that represents a hypothesized relationship between independent and dependent variables, data $D$ is collected to evaluate how well this relationship holds out. A skilled enough model $M$ is retained and used to make predictions; it is put into production. However practical problems arise because under the requentist framework, model evaluation centers on the likelihood $p(D∣M)$, the probability of observing the data given the model is true. With enough data the frequentist assumption is that this likelihood can be directly flipped to give the probability of the model being true, $p(M∣D)$. The leap of faith is completed by imposing unverified assumptions to allow setting model parameters by maximizing $p(D∣M)$. By this logical doing so provides the best model possible. The historical dimensions of how this snake oil was progressively brewed are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

Bayesianism, Frequentism's archrival, provides the framoework to establish the plausibility of $M$ using observations through the reinterpretation of how probability should be construed. As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a statement of our knowledge rather than as an inherent property of the data. the Bayesian perspective treats the data, once observed, as fixed and the model parameters as latent random variables the uncertainty of which  can—and should—be explicitly modeled via prior distributions. Prior distributions are probabilistic constructs that help workers encode prior knowledge. Priors can represents knowledge elicited from subject matter experts, assumptions held by the modeler or a complete absence thereof.

This conceptual framework dissolves the false dichotomy between empirical and semi‑analytical approaches. Both are, in essence, different hypotheses or approximations about how ocean color data are generated; they merely provide alternative expressions of the likelihood 
$p(Chl | Rrs)$ to be updated in light of new data. By adopting a Bayesian paradigm, our approach integrates prior knowledge with observed data to yield a full posterior predictive distribution—offering not only improved predictive performance but also principled uncertainty quantification. In doing so, it circumvents the need for the proliferation of narrowly tuned, sensor-specific variants (65 versions for 25 sensors, as documented by O’Reilly and Werdell [@OREILLY2019]) that are symptomatic of ad hoc calibration processes. Detailed methodological comparisons and additional model refinements are provided in the Supplementary Material.

>>>>

#### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color algorithms relies on a fundamental statistical error. The error is the mistaken assumption that the relative frequencies observed in a finite sample can be taken as precise estimates of the underlying true probabilities. In other words, it is the error of conflating  empirical frequencies — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling. The fallout has and continues to be extremely detrimental. Traditional, frequentist approaches, where the variability in sample frequencies is erroneously treated as negligible, yield overconfident inferences, and allegedgly significant findings that do not hold water, in time leading to replication crises.

In statistical modeling, the goal is to assess how well a hypothesized model $M$ represents the relationship between independent and dependent variables based on collected data $D$. Frequentist methods focus exclusively on the likelihood $p(D|M)$, estimating model parameters by maximizing this probability. This approach assumes that, with a sufficiently large dataset, the chosen parameters reliably reflect the true system. However, in practice, any finite dataset carries inherent uncertainty and noise. As a consequence, models developed under this framework tend to be overconfident---yielding point estimates that do not adequately capture the variability in in situ observations---and may not generalize well to new data. Ad hoc attempts to generate uncertainty, such as bootstrapping, further compound the problem by resampling the flawed, finite observations and thus amplifying any existing biases. This overconfidence, coupled with unverified assumptions presented as objectivity, can lead to misleading inferences and contribute to replication challenges. The historical process by which these assumptions have been enshrined is fascinating in its own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a degree of knowledge rather than as a long-run frequency. Accordingly in this Bayesian framework, the prior $p(M)$ encodes our current state of knowledge about the model's parameters before $D$ is observed. This knowledge can be elicited from experts, or if none exixts weakly informative priors can be used. Note that whether $M$ is based on empirically observed relationships, analytical understanding, or something in between is irrelevant to its integration into this framework. Once the data $D$ are collected, Bayes' theorem combines the likelihood $p(D|M)$ with the prior $p(M)$ to yield  the posterior distribution $p(M|D)$, which represents an updated level of knowledge about the model's parameters. Crucially this posterior can then serve as the new prior when new data become available, allowing for sequential integration of new information. The by-product of this approach is uncertainty estimates that more accurately reflect the complexities of the underlying system as currently understood. In effect, rather than relying on single point estimates based on $p(D|M)$, the posterior $p(M|D)$ can be mined to retrieve a range of plausible parameter values through credible intervals offering a more transparent and robust basis for inference. This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.

In the Bayesian framework, the prior 
p(M) encodes our current state of knowledge about the model's parameters before any data are observed. This knowledge can be elicited from experts or determined experimentally when expert opinion is unavailable. Once data 
D are collected, Bayes’ theorem combines the likelihood 
p(D∣M) with the prior to yield the posterior distribution 
p(M∣D), which represents our updated belief about the parameters. Crucially, this posterior can then serve as the new prior when new data become available, allowing for seamless sequential updating. In this way, whether the model is derived empirically, analytically, or somewhere in between, it is treated as a hypothesis whose parameters are updated in light of new data. This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.
